---
title: "Reproducible Research: Peer Assessment 1"
author: Lihua
output: html_document
keep_md: true
---


## Loading and preprocessing the data
1. code for reading in the dataset and/or processing the data
```{r echo=TRUE}
setwd("D:\\Documents\\gitrepos\\lisa_github\\RepData_PeerAssessment1")
library(dplyr)
library(ggplot2)

dataurl <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip'
# Download Data
if(!file.exists("activity.csv")){
  download.file(dataurl, destfile = 'repdata%2Fdata%2Factivity.zip', mode = "wb")
  unzip(zipfile = 'repdata%2Fdata%2Factivity.zip' )
}

data = read.csv('activity.csv', na.strings = "NA")
data$date = as.Date(data$date, "%Y-%m-%d")
dim(data)
head(data)
```


## What is mean total number of steps taken per day?
2. histogram of the total number of steps taken each day
```{r echo=TRUE, fig.width=6, fig.height=6}
total_steps_per_day = tapply(data$steps, data$date, sum)
total_steps_per_day
hist(total_steps_per_day, xlab = "total steps per day", main = "Total number of steps per day")
```

3. mean and median number of steps taken each day
```{r echo=TRUE}
# mean 
mean(total_steps_per_day, na.rm = TRUE)
# median
median(total_steps_per_day, na.rm = TRUE)
```

## What is the average daily activity pattern?
4. time series plot of the average number of steps taken
```{r echo=TRUE, fig.width=6, fig.height=6}
dim(data)
data_full = data[complete.cases(data),]
dim(data_full)
ave_steps_per_interval = tapply(data_full$steps, data_full$interval, mean)
length_interval = length(unique(data_full$interval))
plot(data_full$interval[1:length_interval], ave_steps_per_interval, type = 'l', xlab = '5 min interval', ylab = 'average steps', main = 'Average steps per 5 min interval accross all days')
```

5. which 5-minute interval that, on average, accorss all the days in the dataset, contains the maximum number of steps
```{r}
the_interval = as.numeric(which.max(ave_steps_per_interval))
data$interval[the_interval]
```

## Imputing missing values
6. code to describe and show a strategy for imputing missing data
```{r echo=TRUE}
# total missing values
print(sum(is.na(data$date)))
print(sum(is.na(data$interval)))
total_missing_values = sum(is.na(data))
print(total_missing_values)

# using the mean value of the 5-min interval to impute the missing values
data_imputed = data

for (i in 1: dim(data)[1]) {
  if(is.na(data$steps[i])) {data_imputed$steps[i]=mean(data[data$interval==data$interval[i],1], na.rm = TRUE)}
}
sum(complete.cases(data_imputed))

total_steps_per_day = tapply(data_imputed$steps, data_imputed$date, sum)

hist(total_steps_per_day, xlab = "total steps per day", main = "Total number of steps per day")

# mean total number of steps taken per day
mean(total_steps_per_day)

# median total number of steps taken per day
median(total_steps_per_day)

# the mean value does not differ from the estimates from the first part of the assignment since we used mean to impute the missing data, but the median value is different from the first part. The impact of imputing missing data on the estimates of the total daily number of steps is that it increases the value of the total daily number of steps.
```


## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
# create a new factor variable in the dataset 
data$weekdayclass <- ifelse(weekdays(as.Date(data$date)) %in% c('Saturday', 'Sunday'), 'weekend', 'weekday')
```

8. panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
```{r echo=TRUE}
interval_mean <- data %>% group_by(interval, weekdayclass) %>% summarise(mean_step = mean(steps, na.rm = TRUE))

fig <- ggplot(interval_mean, aes(interval, mean_step)) + geom_line() + facet_grid(weekdayclass ~ .) + scale_x_continuous("Day Interval", breaks = seq(min(interval_mean$interval), max(interval_mean$interval), 100)) + scale_y_continuous("Average number of steps") + ggtitle("Average number of steps taken by interval")
fig
```



